---
title: 'LocalCop: An R package for local likelihood inference for conditional copulas'
params:
  do_calc: FALSE
tags:
  - R
  - C++
  - copula
  - local likelihood
  - covariate effect
authors:
  - name: Elif Fidan Acar
    orcid: 
    affiliation: "1, 2" # (Multiple affiliations must be quoted)
  - name: Martin Lysy
    orcid: 
    affiliation: 3
  - name: Alan Kuchinsky
    orcid: 
    affiliation: 1
affiliations:
 - name: University of Manitoba
   index: 1
 - name: Hospital for Sick Children
   index: 2
 - name: University of Waterloo
   index: 3
citation_author: Acar et. al.
date: 11 March 2024
year: 2024
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::joss_article
    toc: false
    keep_md: false
csl: apa.csl
journal: JOSS
link-citations: true
---

<!-- !!! ---- WARNING ---- !!!

~~In order to compile this file, you must run `source("joss_article2.R")` first, due to a bug in `rticles::joss_article()`.~~

**Update:** The bug has now been fixed on the development version of [**rticles**](https://github.com/rstudio/rticles).  In order to compile the manuscript, it now suffices to have this version installed: `remotes::install_github('rstudio/rticles', upgrade = TRUE)`.

-->

```{r setup, include = FALSE}
knitr::read_chunk("calculations.R")
library(LocalCop)
library(VineCopula)  # for simulating copula data
library(readr)       # for reading in the copula table
library(kableExtra)  # for styling the copula table
library(tidyr)       # for data manipulations
library(dplyr)       #
library(ggplot2)     # for plots
library(ggpubr)      # 
library(gamCopula)   # other packages for fitting
library(CondCopulas) # conditional copula models
```

<!-- latex macros -->
\newcommand{\R}{{\textsf{R}}}
\newcommand{\cpp}{{\textsf{C++}}}

# Summary

Conditional copulas models allow the dependence structure between multiple response variables to be modelled as a function of covariates.  **LocalCop** is an \R/\cpp package for computationally efficient semiparametric conditional copula modelling using a local likelihood inference framework developed in @ACY2011, @ACY2013 and @ACL2019.

# Statement of Need

There are well-developed \R packages such as **copula** [@copula1; @copula2; @copula3; @copula4] and **VineCopula** [@vinecopula] for fitting copulas in various multivariate data settings. However, these software focus exclusively on unconditional dependence modelling and do not accommodate covariate information. 

Aside from **LocalCop**, \R packages for fitting conditional copulas are **gamCopula** [@gamcopula] and **CondCopulas** [@condcopulas]. **gamCopula** estimates the covariate-dependent copula parameter using spline smoothing.  While this typically has lower variance than the local likelihood estimate provided by **LocalCop**, it also tends to have lower accuracy [@ACL2019].  **CondCopulas** estimates the copula parameter using a semi-parametric maximum-likelihood method based on a kernel-weighted conditional concordance metric.  **LocalCop** also uses kernel weighting, but it uses the full likelihood information of a given copula family rather than just that contained in the concordance metric, and is therefore more statistically efficient.

Local likelihood methods typically involve solving a large number of low-dimensional optimization problems and thus can be computationally intensive.  To address this issue, **LocalCop** implements the local likelihood function in \cpp, using the \R/\cpp package **TMB** [@kristensen.etal16] to efficiently obtain the associated score function using automatic differentiation.  Thus, **LocalCop** is able to solve each optimization problem very quickly using gradient-based algorithms.  It also provides a means of easily parallelizing the optimization across multiple cores, rendering **LocalCop** competitive in terms of speed with other available software for conditional copula estimation.

# Background

For any bivariate response vector $(Y_1, Y_2)$, the conditional joint distribution given a covariate $X$ is given by 
\begin{equation}
F_X(y_1, y_2 \mid x) = C_X (F_{1\mid X} (y_1 \mid x),F_{2\mid X} (y_2 \mid x) \mid x ),
(\#eq:fullmodel)
\end{equation}
where $F_{1\mid X}(y_1 \mid x)$ and $F_{2\mid X}(y_2 \mid x)$ are the conditional marginal distributions of $Y_1$ and $Y_2$ given $X$, and $C_X(u, v \mid x)$ is a conditional copula function.  That is, for given $X = x$, the function $C_X(u, v \mid x)$ is a bivariate CDF with uniform margins. 

The focus of **LocalCop** is on estimating the conditional copula function, which is modelled semi-parametrically as 
\begin{equation}
C_X(u, v \mid x) = \mathcal{C}(u, v\mid \theta(x), \nu),
(\#eq:copmod)
\end{equation}
where $\mathcal{C}(u, v \mid \theta, \nu)$ is one of the parametric copula families listed in Table \@ref(tab:calib), the copula dependence parameter $\theta \in \Theta$ is an arbitrary function of $X$, and $\nu \in \Upsilon$ is an additional copula parameter present in some models. Since most parametric copula families have a restricted range $\Theta \subsetneq \mathbb{R}$, we describe the data generating model (DGM) in terms of the calibration function $\eta(x)$, such that
\begin{equation}
\theta(x) = g^{-1}(\eta(x)),
\end{equation}
where $g^{-1}: \mathbb{R} \to \Theta$ an inverse-link function which ensures that the copula parameter has the correct range. The choice of $g^{-1}(\eta)$ is not unique and depends on the copula family. Table \@ref(tab:calib) displays the copula function $\mathcal{C}(u, v \mid \theta, \nu)$ for each of the copula families provided by **LocalCop**, along with other relevant information including the canonical choice of the inverse link function $g^{-1}(\eta)$.  In Table \@ref(tab:calib), $\Phi^{-1}(p)$ denotes the inverse CDF of the standard normal; $t_{\nu}^{-1}(p)$ denotes the inverse CDF of the Student-t with $\nu$ degrees of freedom; $\Phi_{\theta}(z_1, z_2)$ denotes the CDF of a bivariate normal with mean $(0, 0)$ and variance $\left[\begin{smallmatrix}1 & \theta \\ \theta & 1\end{smallmatrix}\right]$; and $t_{\theta,\nu}(z_1, z_2)$ denotes the CDF of a bivariate Student-t with location $(0, 0)$, scale $\left[\begin{smallmatrix}1 & \theta \\ \theta & 1\end{smallmatrix}\right]$, and degrees of freedom $\nu$.


```{r calib, echo = FALSE}
tab <- read_csv("../copula_table.csv",
                show_col_types = FALSE)
tab_cap <- "Copula families implemented in \\textbf{LocalCop}."
kableExtra::kbl(tab, booktabs = TRUE, escape = FALSE,
                caption = tab_cap) %>%
  kable_styling(latex_options = "scale_down")
```

Local likelihood estimation of the conditional copula parameter $\theta(x)$ uses Taylor expansions to approximate the calibration function $\eta(x)$ at an observed covariate value $X = x$ near a fixed point $X = x_0$, i.e.,
$$
\eta(x)\approx \eta(x_0) + \eta^{(1)}(x_0) (x - x_0) + \ldots + \dfrac{\eta^{(p)}(x_0)}{p!} (x - x_0)^{p}.
$$
One then estimates $\beta_k = \eta^{(k)}(x_0)/k!$ for $k = 0,\ldots,p$ using a kernel-weighted local likelihood function
\begin{equation}
\ell(\boldsymbol{\beta}) = \sum_{i=1}^n \log\left\{ c\left(u_i, v_i \mid g^{-1}( \boldsymbol{x}_{i}^T \boldsymbol{\beta}), \nu \right)\right\} K_h\left(\dfrac{x_i-x_0}{h}\right),
(\#eq:locallik)
\end{equation}
where $(u_i, v_i, x_i)$ is the data for observation $i$, $\boldsymbol{x}_i = (1, x_i - x_0, (x_i - x_0)^2, \ldots, (x_i - x_0)^p)$, $\boldsymbol{\beta}= (\beta_0, \beta_1, \ldots, \beta_p)$, and $K_h(z)$ is a kernel function with bandwidth parameter $h > 0$.  Having maximized $\ell(\boldsymbol{\beta})$ in \@ref(eq:locallik), one estimates $\eta(x_0)$ by $\hat \eta(x_0) = \hat \beta_0$.  Usually, a linear fit with $p=1$ suffices to obtain a good estimate in practice.

# Usage

**LocalCop** is available on [CRAN]() and [GitHub](https://github.com/mlysy/LocalCop).  The two main package functions are:

- `CondiCopLocFit()`: For estimating the calibration function at a sequence of values $\boldsymbol{x}_0 = (x_{01}, \ldots, x_{0m})$.

- `CondiCopSelect()`: For selecting a copula family and bandwidth parameter using leave-one-out cross-validation (LOO-CV) with subsampling as described in @ACL2019.

In the following example, we illustrate the model selection/tuning and fitting steps for data generated from a Clayton copula with conditional Kendall $\tau$ displayed in Figure \@ref(fig:copcomp).  The CV metric for each combination of family and bandwidth are displayed in Figure \@ref(fig:select1-plot).

```{r libs}
```

```{r dgm1-setup}
```
```{r dgm1}
<<dgm>>
```
```{r select1-precalc}
<<select-precalc>>
```
```{r select1-calc, eval = params$do_calc, warning = FALSE}
<<select-calc>>
```
```{r select1-save, include = FALSE, eval = params$do_calc}
saveRDS(cvselect, file = "cvselect1.rds")
```
```{r select1-load, include = FALSE, eval = !params$do_calc}
cvselect <- readRDS("cvselect1.rds")
```
```{r select1-plot, echo = FALSE, fig.height = 2, fig.width = 9, out.width = "100%", fig.cap = "Cross-validation metric for each combination of family and bandwidth."}
<<select-plot>>
```

```{r locfit1}
<<locfit>>
```
```{r gamfit1, include = FALSE}
<<gamfit>>
```
```{r condfit1, include = FALSE}
<<condfit>>
```
```{r sumfit1, include = FALSE}
sumfit1 <- tibble(
  x = x0,
  True = BiCopEta2Tau(family = family, eta = eta_fun(x0)),
  LocalCop = tau_loc,
  gamCopula = tau_gam,
  CondCopulas = tau_cond
)
```
```{r dgm2-setup, include = FALSE}
```
```{r dgm2}
<<dgm>>
```
```{r select2-precalc}
<<select-precalc>>
```
```{r select2-calc, eval = params$do_calc, warning = FALSE}
<<select-calc>>
```
```{r select2-save, include = FALSE, eval = params$do_calc}
saveRDS(cvselect, file = "cvselect2.rds")
```
```{r select2-load, include = FALSE, eval = !params$do_calc}
cvselect <- readRDS("cvselect2.rds")
```
```{r locfit2}
<<locfit>>
```
```{r gamfit2, include = FALSE}
<<gamfit>>
```
```{r condfit2, include = FALSE}
<<condfit>>
```
```{r sumfit2, include = FALSE}
sumfit2 <- tibble(
  x = x0,
  True = BiCopEta2Tau(family = family, eta = eta_fun(x0)),
  LocalCop = tau_loc,
  gamCopula = tau_gam,
  CondCopulas = tau_cond
)
```
```{r copcomp, echo = FALSE, fig.height = 4.5, fig.width = 9, out.width = "100%", fig.cap = "True vs estimated conditional Kendall $\\tau$ using various methods."}
<<plotfit>>

plot1 <- plotfit(sumfit1) + ggtitle("(a) n = 300")
plot2 <- plotfit(sumfit2) + ggtitle("(b) n = 1000")

ggarrange(plot1, plot2, nrow = 2, common.legend = TRUE)
```

In Figure \@ref(fig:copcomp), we compare the true conditional Kendall $\tau$ to estimates using each of the three conditional copula fitting packages **LocalCop**, **gamCopula**, and **CondCopulas**, for sample sizes $n = 300$ and $n = 1000$.  In **gamCopula**, selection of the copula family smoothing splines is done using the generalized CV framework provided by the \R package **mgcv** [@wood17].  In **CondCopulas**, selection of the bandwidth parameter is done using LOO-CV.  In this particular example, the sample size of $n = 300$ is not large enough for **gamCopula** to pick a sufficiently flexible spline basis, and **CondCopulas** picks a large bandwidth which oversmooths the data.  For the larger sample size $n = 1000$, the three methods exhibit similar accuracy.

# Acknowledgements

We acknowledge funding support from the Natural Sciences and Engineering Research Council of Canada Discovery Grants RGPIN-2020-06753 (Acar) and RGPIN-2020-04364 (Lysy).

# References
