---
title: "LocalCop-vignette"
output: 
  bookdown::html_vignette2:
    toc: true
bibliography: references.bib 
vignette: >
  %\VignetteIndexEntry{LocalCop-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This vignette demonstrates the local likelihood inference for the dependence parameters in conditional copula models based on the works @ACY2011, @ACY2013 and @ACL2019.
We consider the bivariate random vector $(Y_1, Y_2)$ whose conditional joint distribution given a covariate $X$ is characterized by 
\begin{equation}
F_X(y_1, y_2 \mid x) = C_X (F_{1\mid X} (y_1 \mid x),F_{2\mid X} (y_2 \mid x) ),
\end{equation}
where $F_{1\mid X}\equiv U$ and $F_{2\mid X}\equiv V$ are conditional marginal distributions of $Y_1$ and $Y_2$ given $X$ and $C_X$ is the conditional copula. 


```{r libs, message=FALSE}
require(LocalCop)
require(VineCopula) # for simulating copula data 
```



## Data generating model (DGM)

We consider parametric copula families $C_X \in \mathcal{C}\{ C(\cdot, \cdot; \theta (X))\}$ to specify the conditional copula. Since most parametric copula families have a restricted parameter range, we describe the data generating model in terms of the calibration function $\eta(x) = g^{-1} (\theta(x))$, where $g^{-1}$ is the inverse link function that maps the copula parameter to $(-\infty, \infty)$. The choice of $g^{-1}$ is not unique and depends on the copula family. The canonical choices for the inverse link function are defined in `BiCopPar2Eta()` for various families (see `ConvertPar()` for details). The results are typically presented in the Kendall tau scale for interpretability across different copula families. 


Below, we provide a walk-through example using the Gumbel family, where the Kendall $\tau$ changing as a function of the covariate in Figure \@ref(fig:dgm). 

```{r dgm, message=FALSE, fig.align='center', fig.width=6, fig.height=6, fig.cap="Conditional Kendall $\\tau$ for Gumbel copula under DGM."}
# simulate copula data given a covariate
set.seed(2024)
family <- 4  # Gumbel Copula 
n <- 1000    # number of observations
X <- sort(runif(n))  # covariate values
eta_fun <- function(x){ # calibration function
 sin(6*pi*x)
}
eta_true <- eta_fun(X)
par_true <- BiCopEta2Par(family = family, eta = eta_true) # copula parameter
udata <- VineCopula::BiCopSim(n, family = family, par = par_true$par)

tau_true <- VineCopula::BiCopPar2Tau(family, par = par_true$par)
plot(X, tau_true, type = "l", lwd=2, ylim=c(-1,1), ylab=expression(tau("x")))

```


For data generation, one can specify conditional marginal models and obtain the response data in the observation scale rather than the uniform scale. This would allow investigating the impact of estimating or misspecifying the conditional marginal models in practical applications. For simplicity, we will proceed with the uniform data assuming that the true marginal models are used. 


## Local likelihood estimation of the dependence parameter

The local likelihood estimation uses the Taylor approximation to approximate the dependence parameter function at an observed covariate around a fixed point $x_0$, i.e.,
$$
\eta(X_i)\approx \eta(x_0) + \eta^{(1)}(x_0) (X_i - x_0) + \ldots + \dfrac{\eta^{(p)}(x_0)}{p!} (X_i - x_0)^{p}.
$$
Usually a linear fit with $p=1$ suffices to obtain a good estimate in practice. 

For a given bandwidth parameter and a copula family, one can obtain the estimates of $\beta_k = \eta^{(k)}(\cdot)/k!$, $k=0,\ldots,p$ using a kernel-weighted local likelihood function
$$
\sum_{i=1}^n \log\{ c(u_i, v_i ; \mathbf{x}_{i,0}^T \boldsymbol{\beta})\} K_h\left(\dfrac{x_i-x_0}{h}\right),
$$
where $\mathbf{x}_{i,0} = (1, x_i-x_0, (x_i-x_0)^2, \ldots, (x_i-x_0)^p)^T$ and $\boldsymbol{\beta}= (\beta_0, \beta_1, \ldots, \beta_p)^T$. This is implemented in `CondiCopLocFit()`.

```{r local1, message=FALSE, fig.align='center', fig.width=6, fig.height=6, fig.cap="Conditional Kendall $\\tau$ for Gumbel copula under DGM."}

x0 <- 0.1
band <- 0.1
fit1 <- CondiCopLocFit(u1 = udata[,1], u2 = udata[,2], family = family, 
                       X = X, x = x0, degree = 1, band = band)
fit1

```

Repeating this procedure over a grid of covariate values allows to capture the underlying dependence parameter function.


```{r localseq, message=FALSE, fig.align='center', fig.width=6, fig.height=6, fig.cap="Conditional Kendall $\\tau$ for Gumbel copula under DGM."}
x0 <- seq(0, 1, by=0.02)
fitseq <- CondiCopLocFit(u1 = udata[,1], u2 = udata[,2], 
               family = family, X = X, x = x0, 
               degree = 1, band = band)

plot(X, tau_true, type = "l", lwd=2,
     ylim=c(-1,1), ylab=expression(tau("x")))
points(x0, BiCopEta2Tau(fitseq$eta, family= family), col=2, pch = 20)

```


## Copula Selection and Model Tuning 


```{r select, message=FALSE, fig.align='center', fig.width=7, fig.height=4, fig.cap="Cross-validated likelihood for copula and bandwidth selection based on a subset."}

bandset <- c(0.1, 0.2, 0.4, 0.8, 1) # bandwidth set
famset <- c(3, 4, 5) # family set
xind <- 100  # number of leave-one-out observations in CV likelihood calculation
system.time({
  cvselect <- CondiCopSelect(u1= udata[,1], u2 = udata[,2],
                          X = X, family = famset, band = bandset,
                          xind = xind)
})
cvselect$cv

cv_res <- cvselect$cv
cv_res <- reshape(cv_res,idvar = "band", timevar = "family", direction = "wide") 
rownames(cv_res) <- paste0("band = ", bandset)
cv_res <- cv_res[,-1]
colnames(cv_res) <- c("Clayton", "Gumbel", "Frank")
cv_res <- as.matrix(cv_res) 
barplot(height =cv_res, beside = TRUE)
opar <- par(oma = c(0,0,0,0), mar = c(0,0,0,0), new = TRUE)
legend(x = "bottom", legend = rownames(cv_res), fill = grey.colors(5),
       bty = "n", ncol = 5, inset = -0.35)
par(opar) # reset par

```


To reduce the computatioanl time in model selection, we calculate the cross-validated likelihood for a subset of the sample. The leave-one-out cross-validation for the full sample is provided below for comparisons. 

```{r select2, message=FALSE, fig.align='center', fig.width=7, fig.height=4, fig.cap="Cross-validated likelihood for copula and bandwidth selection based on the full sample."}

xind <- n 
system.time({
  cvselect <- CondiCopSelect(u1= udata[,1], u2 = udata[,2],
                          X = X, family = famset, band = bandset,
                          xind = xind)
})
cvselect$cv

cv_res <- cvselect$cv
cv_res <- reshape(cv_res,idvar = "band", timevar = "family", direction = "wide") 
rownames(cv_res) <- paste0("band = ", bandset)
cv_res <- cv_res[,-1]
colnames(cv_res) <- c("Clayton", "Gumbel", "Frank")
cv_res <- as.matrix(cv_res) 
barplot(height =cv_res, beside = TRUE)
opar <- par(oma = c(0,0,0,0), mar = c(0,0,0,0), new = TRUE)
legend(x = "bottom", legend = rownames(cv_res), fill = grey.colors(5),
       bty = "n", ncol = 5, inset = -0.35)
par(opar) # reset par

```



## Testing Covariate Effects

