---
title: "LocalCop-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LocalCop-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates local likelihood inference for the dependence parameters in conditional copula models.
We consider the bivariate random vector $(Y_1, Y_2)$ whose joint distribution conditional on the covariate $X$ is characterized by 
\begin{equation}
F_X(y_1, y_2 \mid x) = C_X (F_{1\mid X} (y_1 \mid x),F_{2\mid X} (y_2 \mid x) )
\end{equation}


```{r libs, message=FALSE}
require(LocalCop)
require(VineCopula) # for simulating copula data 
```



## Data generating model (DGM)

The data generating model is described in terms of the calibration function given by $\eta(x) = g^{-1} (\theta(x))$, where $g^{-1}$ is the inverse link function that maps the copula parameter to $(-\infty, \infty)$.


```{r dgm, message=FALSE}

# simulate copula data given a covariate

set.seed(2024)
family <- 5  # Frank Copula 
n <- 1000    # number of observations
X <- runif(n)  # covariate values
eta_fun <- function(x){ # calibration function
 3*sin(6*pi*x)
}
eta_true <- eta_fun(X)
par_true <- BiCopEta2Par(family = family, eta = eta_true) # copula parameter
udata <-  VineCopula::BiCopSim(n, family = family,
                               par = par_true$par)

tau_true <- VineCopula::BiCopPar2Tau(family, 
                                     par = par_true$par)
plot(X, tau_true,
     ylim=c(-1,1), 
     ylab=expression(tau("x")))

```



